---
title: "Assignment 2: Coding Basics"
author: "Candela Cerpa"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.
```{r}
OneToThirthy <- seq(1,30, by=3)
print(OneToThirthy)
```
2.  Compute the mean and median of this sequence.
```{r}
MeanOTT <- mean(OneToThirthy)
print(MeanOTT)
MedianOTT <- median(OneToThirthy)
print(MedianOTT)
```
3.  Ask R to determine whether the mean is greater than the median.
```{r}
Greater <- MeanOTT>MedianOTT
print(Greater)
```
4.  Insert comments in your code to describe what you are doing.
```{r basics 1}
#1. I assigned the sequence a name and applied the sequence function, noting the value by which to increase
#Print shows me the function's result

#2. I assigned each measure a name and applied the corresponding function
#Print shows me the function's result

#3. The function returns FALSE, so mean is not greater than the median
#Print shows me the function's result

```

## Basics, Part 2
5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.
```{r}
Student <- c("Candela", "Samantha", "Elizabeth", "Renata")
TestScore <- c(40, 80, 90, 100)
PassedTest <- TestScore > 50
```
6.  Label each vector with a comment on what type of vector it is.
```{r}
typeof(Student)
#The Student vector is a character type
typeof(TestScore)
#The TestScore vector is a double type
typeof(PassedTest)
#The PassedTest vector is a logical type
```
7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.
```{r}
CourseName <- data.frame(Student, TestScore, PassedTest)
```
8.  Label the columns of your data frame with informative titles.
```{r basics 2}
colnames(CourseName)<-c("Name","Score","Passed")
```

9.  QUESTION: How is this data frame different from a matrix?
> Answer: This dataframe has multiple data classes in its various columns (character, numeric, logical), while a matrix can only have a single type of data class.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 
```{r}
TestCalc1 <- function(x){
  
   for (e in x){
     if (e >= 50) {print(TRUE) } else {print(FALSE)}
   }
    
}

TestCalc2 <- function(x){
  
  for (e in x){
    ifelse(e >= 50, print(TRUE), print(FALSE))
  }
  
}

TestCalc3 <- function(x){
      ifelse(x >= 50, print(TRUE), print(FALSE))
  }

TestCalc4 <- function(x){
     if (x >= 50) {print(TRUE) } else {print(FALSE)}
   }

```
11. Apply your function to the vector with test scores that you created in number 5.
```{r basics 3}
TestCalc1(TestScore)
TestCalc2(TestScore)
TestCalc3(TestScore)
TestCalc4(TestScore)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?
> Answer: if/else would fail without a for loop because it can't apply a logical .
