---
title: "Assignment 5: Data Visualization"
author: "Candela Cerpa"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 load packages, verify directory, read it data
#install.packages("cowplot")
library(tidyverse);library(lubridate);library(here);library(cowplot);library(ggthemes)
getwd()

NTL_LTER <- read.csv("./Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)
NEON_NIWO <- read.csv("./Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_processed.csv", stringsAsFactors = TRUE)

#2 check date formats and adapt
print(class(NTL_LTER$sampledate)) #read as factor
print(class(NEON_NIWO$collectDate)) #read as factor

NTL_LTER$sampledate <- ymd(NTL_LTER$sampledate)
NEON_NIWO$collectDate <- ymd(NEON_NIWO$collectDate)

print(class(NTL_LTER$sampledate)) #read as date
print(class(NEON_NIWO$collectDate)) #read as date
```

## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3 Create a custom theme
my_theme <- theme_base() +
  theme(
    line = element_line(
      linewidth = 2
      ),
    plot.title = element_text(
      color = "darkgreen"
      ),
    axis.ticks = element_line(
      color = "darkgreen"
      ),
    plot.background = element_rect(
      color = "black",
      fill = "grey"
      ),
    complete = TRUE
    )
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 NLT-LTER tp_ug by po4. separate aesthetics for Peter & Paul Lakes

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip:
* Recall the discussion on factors in the previous section as it may be helpful here. 
* R has a built-in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5a boxplot for temp (month on x axis, lake as a color aesthetic)
o5a_plot <- ggplot(NTL_LTER) +
  geom_boxplot(aes(x= month, y = temperature_C))
print(o5a_plot)
#5b boxplot for TP (month on x axis, lake as a color aesthetic)

#5c boxplot for TN (month on x axis, lake as a color aesthetic)

#5d cowplot that combines the 3

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6


#7


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: